---
title: "Basic Competition Aquatic Mosquito Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Competition Aquatic Mosquito Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xDE)
```

Something something.

# Differential Equations

Given $\Lambda$ and some egg laying rate from the adult mosquito population we could formulate and solve a dynamical model of aquatic mosquitoes to give that emergence rate. However, in the example here we will simply use a trace-based (forced) emergence model, so that $\Lambda$ completely specifies the aquatic mosquitoes.

The simplest model of aquatic (immature) mosquito dynamics with negative feedback (density dependence) is:

$$
\dot{L} = \eta - (\psi+\phi+\theta L)L
$$

Because the equations allow the number of larval habitats $l$ to differ from $p$, in general the emergence rate is given by:

$$
\Lambda = \mathcal{N}\cdot \alpha
$$

Where $\mathcal{N}$ is a $p\times l$ matrix and $\alpha$ is a length $l$ column vector given as:

$$
\alpha = \frac{\psi L}{2}
$$

In general, if we know the value of $\Lambda$ at equilibrium we can solve for $L$ directly by using the above two equations. Then we can consider $\theta$, the strength of density dependence to be unknown and solve such that:

$$
\theta = (\eta - \psi L - \phi L) / L^2
$$
