---
title: "Trace (exogenously forced) Aquatic Mosquito Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trace (exogenously forced) Aquatic Mosquito Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is the null model of aquatic mosquito dynamics; there are no endogenous dynamics
and the model is simply specified by `Lambda`, the rate at which adult mosquitoes
spontaneously emerge from aquatic habitats.

Below we show an example, taken from the package tests, of using the trace-based
aquatic model to keep an adult mosquito model at equilibrium when using unequal numbers
of aquatic habitats per patch.

We first calculate equilibrium values following the [Ross-Macdonald vignette](adult_RM.html).
We use $\mathcal{N}$ and $\mathcal{U}$ to describe how aquatic habitats are dispersed amongst
patches, and how mosquitoes in each patch disperse eggs to each habitat, respectively.
Please note because we have unequal numbers of aquatic habitats and patches, we use `MASS::ginv`
to compute the generalized inverse of $\mathcal{N}$ to get $\alpha$ required at equilibrium.

```{r}
library(exDE)
library(MASS)
library(expm)
library(deSolve)
library(data.table)
library(ggplot2)

nPatches <- 3
nHabitats <- 4
f <- 0.3
q <- 0.9
g <- 1/20
sigma <- 1/10
tau <- 11
nu <- 1/2
eggsPerBatch <- 30

calN <- matrix(0, nPatches, nHabitats)
calN[1,1] <- 1
calN[2,2] <- 1
calN[3,3] <- 1
calN[3,4] <- 1

calU <- matrix(0, nHabitats, nPatches)
calU[1,1] <- 1
calU[2,2] <- 1
calU[3:4,3] <- 0.5

calK <- matrix(0, nPatches, nPatches)
calK[1, 2:3] <- c(0.2, 0.8)
calK[2, c(1,3)] <- c(0.5, 0.5)
calK[3, 1:2] <- c(0.7, 0.3)
calK <- t(calK)

Omega <- make_Omega(g, sigma, calK, nPatches)
Upsilon <- expm::expm(-Omega * tau)

kappa <- c(0.1, 0.075, 0.025)
Lambda <- c(5, 10, 8)

# equilibrium solutions
Omega_inv <- solve(Omega)

M_eq <- as.vector(Omega_inv %*% Lambda)
G_eq <- as.vector(solve(diag(nu+f, nPatches) + Omega) %*% diag(f, nPatches) %*% M_eq)
Y_eq <- as.vector(solve(diag(f*q*kappa) + Omega) %*% diag(f*q*kappa) %*% M_eq)
Z_eq <- as.vector(Omega_inv %*% Upsilon %*% diag(f*q*kappa) %*% (M_eq - Y_eq))

# the "Lambda" for the dLdt model
alpha <- as.vector(ginv(calN) %*% Lambda)
```

Now we set up the model. Please see the [Ross-Macdonald vignette](adult_RM.html) for details
on the adult model. We use `alpha` as the `Lambda` parameter of our forced emergence model, in
`exDE::make_parameters_L_trace`. Again, because we are not running the full transmission model,
we must use `exDE::MosquitoBehavior` to pass the equilibrium values of those bionomic parameters 
to `exDE::xDE_diffeqn_mosy`.

```{r, out.width = "100%"}
params <- list(
  nPatches = nPatches,
  nHabitats = nHabitats,
  calU = calU,
  calN = calN
)
params <- list2env(params)

# ODE
make_parameters_MYZ_RM_ode(pars = params, g = g, sigma = sigma, calK = calK, tau = tau, f = f, q = q, nu = nu, eggsPerBatch = eggsPerBatch, M0 = rep(0, nPatches), G0 = rep(0, nPatches), Y0 = rep(0, nPatches), Z0 = rep(0, nPatches))
make_parameters_L_trace(pars = params, Lambda = alpha)
make_indices(params)

y0 <- rep(0, max(params$Upsilon_ix))
y0[params$M_ix] <- M_eq
y0[params$G_ix] <- G_eq
y0[params$Y_ix] <- Y_eq
y0[params$Z_ix] <- Z_eq
y0[params$Upsilon_ix] <- as.vector(Upsilon)

MosyBehavior <- MosquitoBehavior(0, y, params)

out <- deSolve::ode(y = y0, times = 0:50, func = xDE_diffeqn_mosy, parms = params, method = 'lsoda', kappa = kappa, MosyBehavior = MosyBehavior)

colnames(out)[params$M_ix+1] <- paste0('M_', 1:params$nPatches)
colnames(out)[params$G_ix+1] <- paste0('G_', 1:params$nPatches)
colnames(out)[params$Y_ix+1] <- paste0('Y_', 1:params$nPatches)
colnames(out)[params$Z_ix+1] <- paste0('Z_', 1:params$nPatches)
out <- out[, -c(params$Upsilon_ix+1)]

out <- as.data.table(out)
out <- melt(out, id.vars = 'time')
out[, c("Component", "Patch") := tstrsplit(variable, '_', fixed = TRUE)]
out[, variable := NULL]

ggplot(data = out, mapping = aes(x = time, y = value, color = Patch)) +
  geom_line() +
  facet_wrap(. ~ Component, scales = 'free') +
  theme_bw()
```
