---
title: "Basic Competition Aquatic Mosquito Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Competition Aquatic Mosquito Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The basic competition aquatic mosquito model fulfills the generic interface of the aquatic mosquito component. It has a single compartment "larvae" for each aquatic habitat, and mosquitoes in that aquatic habitat suffer density-independent and dependent mortality, and mature at some rate $\psi$.

# Differential Equations

Given $\Lambda$ and some egg laying rate from the adult mosquito population we could formulate and solve a dynamical model of aquatic mosquitoes to give that emergence rate. However, in the example here we will simply use a trace-based (forced) emergence model, so that $\Lambda$ completely specifies the aquatic mosquitoes.

The simplest model of aquatic (immature) mosquito dynamics with negative feedback (density dependence) is:

$$
\dot{L} = \eta - (\psi+\phi+\theta L)L
$$

Because the equations allow the number of larval habitats $l$ to differ from $p$, in general the emergence rate is given by:

$$
\Lambda = \mathcal{N}\cdot \alpha
$$

Where $\mathcal{N}$ is a $p\times l$ matrix and $\alpha$ is a length $l$ column vector given as:

$$
\alpha = \psi L
$$

# Equilibrium solutions

In general, if we know the value of $\Lambda$ at equilibrium we can solve for $L$ directly by using the above two equations. Then we can consider $\theta$, the strength of density dependence to be unknown and solve such that:

$$
\theta = (\eta - \psi L - \phi L) / L^2
$$

# Example

```{r, message=FALSE, warning=FALSE}
library(exDE)
#source ("../R/aquatic-interface.R")
#source ("../R/aquatic-basic.R")
library(deSolve)
library(data.table)
library(ggplot2)
```

Here we run a simple example with 3 aquatic habitats at equilibrium. We use `exDE::make_parameters_L_basic` to
set up parameters. Please note that this only runs the aquatic mosquito component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

```{r, out.width = "100%"}
nHabitats <- 3
alpha <- c(10, 50, 20)
eta <- c(250, 500, 170)
psi <- 1/10
phi <- 1/12

L <- alpha/psi
theta <- (eta - psi*L - phi*L)/(L^2)

params <- list(
  nHabitats = nHabitats
)

params = make_parameters_L_basic(pars = params, psi = psi, phi = phi, theta = theta)
params = make_inits_L_basic(params, L)
params = make_indices(params)

y0 <- get_inits_L(params) 

dLdt_local = function(t, y, pars, eta) {
  list(dLdt(t, y, pars, eta))
}
  

out <- deSolve::ode(y = y0, times = 0:50, dLdt_local, parms = params, method = 'lsoda', eta = eta)

colnames(out)[params$L_ix+1] <- paste0('L_', 1:params$nHabitats)

out <- as.data.table(out)
out <- melt(out, id.vars = 'time')
out[, c("Component", "Patch") := tstrsplit(variable, '_', fixed = TRUE)]
out[, variable := NULL]

ggplot(data = out, mapping = aes(x = time, y = value, color = Patch)) +
  geom_line() +
  theme_bw()
```
