---
title: "SIS (Susceptible-Infected-Susceptible) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIS (Susceptible-Infected-Susceptible) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SIS (Susceptible-Infected-Susceptible) human model model fulfills the generic interface of the human population component. It is the simplest model of endemic diseases in humans.

# Differential Equations

Let the generalized SIS human model be:

$$
\dot{X} = \mbox{diag}(bEIR) \cdot (H-X) - rX
$$

# Equilibrium Solutions

A typical situation when using this model is that $H$ (total population size by strata) and $X$ (number of infectious persons by strata) are known from census and survey data. Then it is of interest to find the value of $EIR$ (Entomological Inoculation Rate) which leads to that prevalence at equilibrium.

$$
0 = \mbox{diag}(bEIR) \cdot (H-X) - rX 
$$

$$ 
rX = \mbox{diag}(b) \cdot \mbox{diag}(EIR) \cdot (H-X) 
$$

$$
\frac{rX}{H-X} = \mbox{diag}(b) \cdot \mbox{diag}(EIR)
$$

$$
\mbox{diag}(1/b) \cdot \left(\frac{rX}{H-X}\right) = EIR
$$

Note that in the final line, $EIR$ is a column vector of dimension $n$ due to the operations on the left. Each element gives the per-capita rate at which individuals in that population strata receive potentially infectious bites (summing across all the places they visit).

# Example

```{r, message=FALSE, warning=FALSE}
library(exDE)
library(deSolve)
library(data.table)
library(ggplot2)
```

Here we run a simple example with 3 population strata at equilibrium. We use `exDE::make_parameters_X_SIS` to
set up parameters. Please note that this only runs the human population component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

We use the null (constant) model of human demography ($H$ constant for all time).

```{r, out.width = "100%"}
nStrata <- 3
H <- c(100, 500, 250)
membershipH <- rep(1,3) 
searchWtsH <- rep(1,3) 
X <- c(20, 120, 80)
b <- 0.55
c <- 0.15
r <- 1/200
Psi <- matrix(data = 1,nrow = 1, ncol = nStrata)

EIR <- diag(1/b, nStrata) %*% ((r*X)/(H-X))

params <- make_parameters_xde()
params$nStrata = nStrata
params$nPatches = 1 

params = make_parameters_demography_static(pars = params, H=H, membershipH=membershipH, searchWtsH=searchWtsH, TimeSpent=Psi, birthF = "constant", birthrate = 0, Hmatrix = diag(1, length(H)), birthsXstrata = rep(1,length(H)))

params = make_parameters_X_SIS(pars = params, b = b, c = c, r = r)

params = make_inits_X_SIS(pars = params, X)
params = make_indices(params)

y0 <- get_inits_X(params)

dXdt_local = function(t, y, pars, EIR) {
  list(dXdt(t, y, pars, EIR))
}

out <- deSolve::ode(y = y0, times = c(0, 365), dXdt_local, parms = params, method = 'lsoda', EIR = as.vector(EIR))

colnames(out)[params$X_ix+1] <- paste0('X_', 1:params$nStrata)

out <- as.data.table(out)
out <- melt(out, id.vars = 'time')
out[, c("Component", "Strata") := tstrsplit(variable, '_', fixed = TRUE)]
out[, variable := NULL]

ggplot(data = out, mapping = aes(x = time, y = value, color = Strata)) +
  geom_line() +
  theme_bw()
```
