---
title: "SIS (Susceptible-Infected-Susceptible) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIS (Susceptible-Infected-Susceptible) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xDE)
```

Let the generalized SIS human model be:

$$
\dot{X} = \mbox{diag}(bEIR) \cdot (H-X) - rX
$$

A typical situation when using this model is that $H$ (total population size by strata) and $X$ (number of infectious persons by strata) are known from census and survey data. Then it is of interest to find the value of $EIR$ (Entomological Inoculation Rate) which leads to that prevalence at equilibrium.

$$
0 = \mbox{diag}(bEIR) \cdot (H-X) - rX 
$$

$$ 
rX = \mbox{diag}(b) \cdot \mbox{diag}(EIR) \cdot (H-X) 
$$

$$
\frac{rX}{H-X} = \mbox{diag}(b) \cdot \mbox{diag}(EIR)
$$

$$
\mbox{diag}(1/b) \cdot \left(\frac{rX}{H-X}\right) = EIR
$$

Note that in the final line, $EIR$ is a column vector of dimension $n$ due to the operations on the left. Each element gives the per-capita rate at which individuals in that population strata receive potentially infectious bites (summing across all the places they visit).
