---
title: "Hybrid MoI (Multiplicity of Infection) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrid MoI (Multiplicity of Infection) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is a hybrid model which tracks the mean multiplicity of infection (superinfection) in two compartments. The first, $m1$ is all infections, and the second $m2$ are apparent (patent) infections. Therefore $m2$ is "nested" within $m1$. It is a "hybrid" model in the sense of [Nåsell (1985)](https://link.springer.com/book/10.1007/978-3-662-01609-1).

# Differential Equations

The equations are as follows:

$$
\dot{m_{1}} = h - r_{1}m_{1}
$$
$$
\dot{m_{2}} = h - r_{2}m_{2}
$$
Where $h = b EIR$, is the force of infection. Prevalence can be calculated from these MoI values by:

$$
x_{1} = 1-e^{-m_{1}}
$$
$$
x_{2} = 1-e^{-m_{2}}
$$
The net infectious probability to mosquitoes is therefore given by:

$$
x = c_{2}x_{2} + c_{1}(x_{1}-x_{2})
$$

Where $c_{1}$ is the infectiousness of inapparent infections, and $c_{2}$ is the infectiousness of patent infections.

# Equilibrium solutions

One way to proceed is assume that $m_{2}$ is known, as it models the MoI of patent (observable) infections. Then we have:

$$
h = r_{2}/m_{2}
$$
$$
m_{1} = h/r_{1}
$$
We can use this to calculate the net infectious probability, and then $\kappa = \beta^{\top} \cdot x$, allowing the equilibrium solutions of this model to feed into the other components.

# Example

```{r, message=FALSE, warning=FALSE}
library(exDE)
library(deSolve)
library(data.table)
library(ggplot2)
```

Here we run a simple example with 3 population strata at equilibrium. We use `exDE::make_parameters_X_hMoI` to
set up parameters. Please note that this only runs the human population component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

We use the null (constant) model of human demography ($H$ constant for all time). While $H$ does not appear in the equations above, it must be specified as the model of bloodfeeding ($\beta$) relies on $H$ to compute consistent values.

```{r, out.width = "100%"}
nStrata <- 3
H <- c(100, 500, 250)
membershipH = rep(1,3) 
searchWtsH = rep(1,3) 
b <- 0.55
c1 <- 0.05
c2 <- 0.25
r1 <- 1/250
r2 <- 1/50
Psi <- matrix(data = 1, nrow = 1, ncol = nStrata)

m20 <- 1.5
h <- r2*m20
m10 <- h/r1

EIR <- h/b

params <- make_parameters_xde()
params$nStrata = nStrata
params$nPatches = 1 

params = make_parameters_demography_null(pars = params, H=H, membershipH=membershipH, searchWtsH=searchWtsH, TimeSpent=Psi) 

params = make_parameters_X_hMoI(pars = params, b = b, c1 = c1, c2 = c2, r1 = r1, r2 = r2)
params = make_inits_X_hMoI(pars = params, rep(m10, nStrata), rep(m20, nStrata)) 
params = make_indices(params)

y0 <- get_inits(params)

dXdt_local = function(t, y, pars, EIR) {
  list(dXdt(t, y, pars, EIR))
}

out <- deSolve::ode(y = y0, times = c(0, 365), func = dXdt_local, 
                    parms = params, method = 'lsoda', EIR = as.vector(EIR))

colnames(out)[params$m1_ix+1] <- paste0('m1_', 1:params$nStrata)
colnames(out)[params$m2_ix+1] <- paste0('m2_', 1:params$nStrata)

out <- as.data.table(out)
out <- melt(out, id.vars = 'time')
out[, c("Component", "Strata") := tstrsplit(variable, '_', fixed = TRUE)]
out[, variable := NULL]

ggplot(data = out, mapping = aes(x = time, y = value, color = Strata)) +
  geom_line() +
  facet_wrap(Strata ~ Component, scales = 'free') +
  theme_bw()
```

