---
title: "SIS (Susceptible-Infected-Susceptible) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIS (Susceptible-Infected-Susceptible) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SIS (Susceptible-Infected-Susceptible) human model model fulfills the generic interface of the human population component. It is the simplest model of endemic diseases in humans.

# Differential Equations

We subdivide a population into susceptible ($S$) and infected and infectious ($I$) individuals, where the total population is $H = S+I.$ We assume the force of infection ($h$, FoI) is linearly proportional to the EIR: $h = b \times EIR.$ In its general form, with births ($B(H)$) and deaths (at the per-capita rate $\mu$), the generalized SIS dynamics are:

$$
\begin{array}{rl}
\dot{S} &= -h S + rI + B(H) -\mu S\\
\dot{I} &= h S  - rI - \mu I
\end{array}
$$

If there is no demographic change, the SIS model can be rewritten as a single equation: 

$$
\dot{I} = h (H-I) - rI
$$
Even in this simplified form, we are assuming that a population could be stratified, such that the variables and parameter are all vectors with length `nStrata.`

# Equilibrium Solutions

A typical situation when using this model is that $H$ (total population size by strata) and $X$ (number of infectious persons by strata) are known from census and survey data. Then it is of interest to find the value of $EIR$ (Entomological Inoculation Rate) which leads to that prevalence at equilibrium.

$$
0 = \mbox{diag}(bEIR) \cdot (H-I) - rI 
$$

$$ 
rI = \mbox{diag}(b) \cdot \mbox{diag}(EIR) \cdot (H-I) 
$$

$$
\frac{rI}{H-I} = \mbox{diag}(b) \cdot \mbox{diag}(EIR)
$$

$$
\mbox{diag}(1/b) \cdot \left(\frac{rI}{H-I}\right) = EIR
$$

Note that in the final line, $EIR$ is a column vector of dimension $n$ due to the operations on the left. Each element gives the per-capita rate at which individuals in that population strata receive potentially infectious bites (summing across all the places they visit).

# Example

```{r, message=FALSE, warning=FALSE}
library(exDE)
library(deSolve)
library(data.table)
library(ggplot2)
```

Here we run a simple example with 3 population strata at equilibrium. We use `exDE::make_parameters_X_SIS` to
set up parameters. Please note that this only runs the human population component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

We use the null (constant) model of human demography ($H$ constant for all time).

## The Long Way

To set up systems of differential equations, we must set the values of all our parameters. 

```{r}
nStrata <- 3
H <- c(100, 500, 250)
residence <- rep(1,3) 
searchWtsH <- rep(1,3) 
I <- c(20, 120, 80)
b <- rep(0.55, nStrata) 
c <- rep(0.15, nStrata) 
r <- rep(1/200, nStrata) 
TaR <- matrix(data = 1,nrow = 1, ncol = nStrata)

EIR <- diag(1/b, nStrata) %*% ((r*I)/(H-I))
```


```{r} 
params <- make_parameters_xde()
params$nStrata = nStrata
params$nPatches = 1
params$nHosts = 1 
params$eir = EIR
params$EIR = list() 

params$eir = as.vector(EIR)
F_eir = function(t, pars){
  pars$EIR[[1]] = pars$eir
}

params = make_parameters_demography_null(pars = params, 
              H=H, residence=residence, searchWts=searchWtsH, TaR=TaR)

params = make_parameters_X_SIS(pars = params, b = b, c = c, r = r)

params = make_inits_X_SIS(pars = params, H-I, I)

params = make_indices(params)
```

```{r}
y0 <- get_inits(params)

out <- deSolve::ode(y = y0, times = c(0, 365), xDE_diffeqn_cohort, 
                    parms = params, method = 'lsoda', F_eir = F_eir) 
out1 <- out
```

```{r, out.width = "100%"}
colnames(out)[params$ix$X$S_ix+1] <- paste0('S_', 1:params$nStrata)
colnames(out)[params$ix$X$I_ix+1] <- paste0('I_', 1:params$nStrata)

out <- as.data.table(out)
out <- melt(out, id.vars = 'time')
out[, c("Component", "Strata") := tstrsplit(variable, '_', fixed = TRUE)]
out[, variable := NULL]

ggplot(data = out, mapping = aes(x = time, y = value, color = Strata)) +
  geom_line() +
  facet_wrap(. ~ Component, scales = 'free') +
  theme_bw()

```

## Using Setup

We have developed utilities for setting up models. We pass the parameter values and initial values as lists: 

```{r}
Xo = list(b=0.55, c=0.15, r=1/200, I0 = I)
Hpop = c(100, 500, 250)
```


To use `xde_setup_cohort` we must set up a function `F_eir` returns the value of EIR we computed above:  

```{r}
fF_eir1 = function(EIR){return(function(t, pars){EIR})} 
F_eir1 = fF_eir1(EIR)
```

```{r}
xde_setup_cohort(F_eir1, Xname="SIS", HPop=Hpop, Xopts = Xo) -> test_SIS
```

```{r}
xde_solve(test_SIS, 365, 365)$outputs$orbits$deout -> out2
approx_equal(out2,out1)
```
