---
title: "SIP (Susceptible-Infected-Prophylaxis) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIP (Susceptible-Infected-Prophylaxis) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xDE)
```

The modular nature of the equations makes it relatively simple to add complexity to specific components of the framework without reconfiguring the entire system.

One reasonable complication would be to add treatment followed by a duration of protective immunity to reinfection to the SIS model, creating SIP (Susceptible-Infectious-Protected) dynamics. This requires two new parameters, $\rho$, the probability a new infection is treated, and $\eta$ the duration of chemoprophylaxis following treatment. $X$ remains a column vector giving the number of infectious individuals in each strata, and $P$ the number of treated and protected individuals.

The equations are as follows:

$$
\dot{X} = \mbox{diag}((1-\rho)bEIR)\cdot (H-X-P) - rX
$$

$$
\dot{P} = \mbox{diag}(\rho b EIR) \cdot (H-X-P) - \eta P
$$

Again, we assume $H$ and $X$ to be known. and solve for $EIR$ and $P$.

$$
P = \mbox{diag}(1/\eta) \cdot \mbox{diag}(\rho/(1-\rho)) \cdot rX
$$

$$
EIR = \mbox{diag}(1/b) \cdot \mbox{diag}(1/(1-\rho)) \cdot \left( \frac{rX}{H-X-P} \right)
$$

Given $EIR$ we can solve for the mosquito population which would have given rise to those equilibrium values.
